---
title: "Blood donor recruitment prediction"
author: "Timo Asikainen & Mikko Arvas"
date: "`r Sys.time()`"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)  
library(patchwork)
library(binom)
```

GitHub: https://github.com/FRCBS/donor-recruitment-prediction/blob/main/src/blood-donor-recruitment-exploratory.Rmd


First run read-data-files.Rmd to produce the data files
et.Rdata, et.noage.Rdata, activity.stats.Rdata, activity.stats.sex.Rdata

TODO:


- more on O- data


# Parameters

```{r working-directory-and-file-paths}
param=list()
param$wd = getwd()

# getwd() might behave differently depending on the environment where it is run (console vs. Rmd),
# therefore checking if working directory is set to the src folder and moving up if yes.
if (grepl('[/\\]src[/\\]?',param$wd)) {
   param$wd = sub('[/\\]src([/\\]?)$','\\1',param$wd)
}

dir.create(file.path(param$wd,"results"),showWarnings = FALSE)
dir.create(file.path(param$wd,"log"),showWarnings = FALSE)

# nb! check that the result and log files point to the right locations
# No log is currently written in this script
```


```{r parameters} 

param$wd = getwd()
if (grepl('[/\\]src[/\\]?',param$wd)) {
   param$wd = sub('[/\\]src(data)?([/\\]?)$','\\2',param$wd)
}

param$data.file = file.path(param$wd,'/results/et.Rdata')
# param$data.file = "c:/git_repot/DATA/donationdata.fortimo.rdata" # fi

```

# Load and summarise data

## Donation counts

```{r load-data} 

load(param$data.file)
et <- et %>%
   mutate(
  age.group = str_c(age.lower,"-",age.upper)
) %>% 
  mutate(
  across(where(is.character),as.factor)
) %>% filter(!is.na(cdon))

summary(et)

```

```{r fake-data}

if (TRUE) {
set.seed(2030)
et <- et %>% filter(country == 'fi') %>% 
  mutate(
    country = "FAKE_fo",
    year0 = year0 -1,
    year = year -1,
    cdon = jitter(cdon+1),
    don = jitter(don+1),
    avage = jitter(avage)
  ) %>% bind_rows(et)
et <- et %>% filter(country == 'nl') %>% 
  mutate(
    country = "FAKE_no",
    year0 = year0 -1,
    year = year -1,
    cdon = jitter(cdon+1),
    don = jitter(don+1),
    avage = jitter(avage)
  ) %>% bind_rows(et) %>%  
  mutate(
  across(where(is.character),as.factor)
)
  
summary(et)
}

```

```{r}
et %>% group_by(country,age.lower,age.upper,BloodGroup) %>% count() %>% filter(age.upper == 100 & age.lower == 0)
```

Comparison groups for O- are not there yet.


```{r}
et %>% group_by(country,age.lower,age.upper,BloodGroup) %>% count() %>% filter(!(age.upper == 100 & age.lower == 0))
```



```{r}

et %>% 
  group_by(country,BloodGroup) %>% 
  summarise(
    n = n(),
    meancdon = mean(cdon, na.rm=TRUE),
    meandon = mean(don, na.rm=TRUE),
    .groups = "keep"
  )
  

```

Mean donations by coutry and blood group.

```{r}
# Remove the first year of each country
miny <- et %>% group_by(country) %>% 
  summarise(miny=min(year0))

for(i in miny$country) {
  et <- et %>% 
    filter(! (country == i & year0 == miny[miny$country== i ,]$miny) )
}

# Remove last year from Finnish data as it is not complete
  et <- et %>% 
    filter(!(country == "fi" & year0 == 2024 ))


```


## Probability of next donation 

```{r}
param$data.file = file.path(param$wd,'/results/activity.stats.sex.rdata')
load(param$data.file)

activity.stats.sex <-   activity.stats.sex %>% 
  na.omit() %>% 
  mutate(
    x = round(n * prop,0),
    ciplower =  binom.confint( x = x ,n=n, method=c("ac"),type="central")$lower,
    cipupper =  binom.confint( x = x ,n=n, method=c("ac"),type="central")$upper,
    cidelower = 0.5 * qchisq(0.95/2, 2*delay),
    cideupper = 0.5 * qchisq(1-0.95/2, 2*delay+2)
   ) %>% 
 mutate(
  across(where(is.character),as.factor)
  )
summary(activity.stats.sex)

```


```{r fake-data-2}

if (TRUE) {
set.seed(2030)
activity.stats.sex <- activity.stats.sex %>% filter(country == 'fi') %>% 
  mutate(
    country = "FAKE_fo",
    n = jitter(n+1),
    prop = jitter(prop+0.01),
    delay = jitter(delay+10),
    ciplower = jitter(ciplower+0.01),
    cipupper = jitter(cipupper+0.01),
    cidelower = jitter(cidelower+10),
    cideupper = jitter(cideupper+10),
  ) %>% bind_rows(activity.stats.sex)
activity.stats.sex <- activity.stats.sex %>% filter(country == 'nl') %>% 
  mutate(
    country = "FAKE_nl",
    n = jitter(n+1),
    prop = jitter(prop+0.01),
    delay = jitter(delay+10),
    ciplower = jitter(ciplower+0.01),
    cipupper = jitter(cipupper+0.01),
    cidelower = jitter(cidelower+10),
    cideupper = jitter(cideupper+10)
  ) %>% bind_rows(activity.stats.sex) %>% 
  mutate(
  across(where(is.character),as.factor)
)
summary(activity.stats.sex)
}

```



# Donation counts plots

## Cumulative donation counts raw data

### -O-

```{r}

plotcdon <- function(data){
p <- ggplot(
  data,
  aes(x = year, y = cdon, color = year0, group=year0)
) 
p <- p + geom_path()
p <- p +facet_grid(country~age.group)
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1))
p <- p + ylab("Cumulative count of donations") + xlab("Year")
#p <- p + guides(color = guide_legend(title = "First year")) 
    
}

```


```{r , warning=FALSE, fig.width=12,fig.height=14}

pmo <- plotcdon(et %>% filter(BloodGroup == "-O-" & Sex== "Male", DonationPlaceType == 'Office')) + ggtitle("Male Office")
pfo <- plotcdon(et %>% filter(BloodGroup == "-O-" & Sex== "Female", DonationPlaceType == 'Office')) + ggtitle("Female Office")
pmm <- plotcdon(et %>% filter(BloodGroup == "-O-" & Sex== "Male", DonationPlaceType == 'Mobile')) + ggtitle("Male Mobile")
pfm <- plotcdon(et %>% filter(BloodGroup == "-O-" & Sex== "Female", DonationPlaceType == 'Mobile')) + ggtitle("Female Mobile")

wrap_plots(pmo + pfo + pmm + pfm) + plot_layout(guides = "collect") & 
  scale_colour_continuous(limits = range(et$year0))

```

### O-

```{r , warning=FALSE, fig.width=12,fig.height=14}

pmo <- plotcdon(et %>% filter(BloodGroup == "O-" & Sex== "Male", DonationPlaceType == 'Office')) + ggtitle("Male Office")
pfo <- plotcdon(et %>% filter(BloodGroup == "O-" & Sex== "Female", DonationPlaceType == 'Office')) + ggtitle("Female Office")
pmm <- plotcdon(et %>% filter(BloodGroup == "O-" & Sex== "Male", DonationPlaceType == 'Mobile')) + ggtitle("Male Mobile")
pfm <- plotcdon(et %>% filter(BloodGroup == "O-" & Sex== "Female", DonationPlaceType == 'Mobile')) + ggtitle("Female Mobile")

wrap_plots(pmo + pfo + pmm + pfm) + plot_layout(guides = "collect") & 
  scale_colour_continuous(limits = range(et$year0))

```



## Age and donation count


```{r}
plotaveagecdon <- function(data){
p <- ggplot(
  data,
  aes(x = avage, y = cdon, color = year0,shape=age.group)
) 
p <- p + geom_point()
p <- p + facet_grid(country~DonationPlaceType)
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1))
p <- p + ylab("Cumulative count of donations") + xlab("Average age")
#p <- p + guides(color = guide_legend(title = "First year")) 
    
}


```

```{r , warning=FALSE, fig.width=12,fig.height=14}

pmo <- plotaveagecdon(et %>% filter(BloodGroup == "-O-" & Sex== "Male")) + ggtitle("Male")
pfo <- plotaveagecdon(et %>% filter(BloodGroup == "-O-" & Sex== "Female")) + ggtitle("Female")

wrap_plots(pmo + pfo ) + plot_layout(guides = "collect") 
  

```

## Cumulative donation count country wise comparison


```{r}
plotaveagecdoncountry <- function(data){
p <- ggplot(
  data,
  aes(x = avage, y = cdon, color = country, alpha=rev(year0))
  
) 
#p <- p + geom_point(alpha =0.5)
p <- p + geom_point()
p <- p + facet_grid(DonationPlaceType~.)
p <- p + theme(axis.text.x=element_text(angle=90, hjust=1))
p <- p + ylab("Cumulative count of donations") + xlab("Average age")
p <- p + guides(
  color = guide_legend(title = "Country"),
  alpha = guide_legend(title = "First year")
  ) 
    
}


```

```{r , warning=FALSE, fig.width=12,fig.height=8}

pmo <- plotaveagecdoncountry(et %>% filter(BloodGroup == "-O-" & Sex== "Male")) + ggtitle("Male")
pfo <- plotaveagecdoncountry(et %>% filter(BloodGroup == "-O-" & Sex== "Female")) + ggtitle("Female")

wrap_plots(pmo + pfo ) + plot_layout(guides = "collect") 
  

```


# Next donation plots

## Probability of next donation

```{r}
p <- ggplot(activity.stats.sex)
p <- p + geom_ribbon(aes(x = ord, ymin = ciplower,ymax=cipupper,fill=country), alpha = 0.5)
p <- p + facet_grid(.~Sex)
p <- p + xlab("Number of donation event") + ylab("Probability of next donation")
p
```


```{r}
p <- ggplot(activity.stats.sex)
p <- p + geom_ribbon(aes(x = ord, ymin = ciplower,ymax=cipupper,fill=country), alpha = 0.5)
p <- p + xlim(0,40) # how to take into account variable length of data by country?
p <- p + facet_grid(.~Sex)
p <- p + xlab("Number of donation event") + ylab("Probability of next donation")
p <- p + guides(
  color = guide_legend(title = "Country")
  )
p
```


## Delay to next donation

```{r}
p <- ggplot(activity.stats.sex)
p <- p + geom_ribbon(aes(x = ord, ymin = cidelower,ymax=cideupper,fill=country))
p <- p + facet_grid(.~Sex)
p <- p + scale_y_log10()
p <- p + xlab("Number of donation event") +  ylab("Delay in days to next donation")
p
```


```{r}
p <- ggplot(activity.stats.sex)
p <- p + geom_ribbon(aes(x = ord, ymin = cidelower,ymax=cideupper,fill=country))
p <- p + facet_grid(.~Sex)
p <- p + scale_y_log10()
p <- p + xlim(0,40)  # how to take into account variable length of data by country?
p <- p + xlab("Number of donation event") +  ylab("Delay in days to next donation")
p <- p + guides(
  color = guide_legend(title = "Country")
  )
p
```
